Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-02-15 RNARedPrint-Bioinfo2018.tex
Cosmetic changes

\abstract{\textbf{Motivation:} The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches largely benefit from effective in-silico methods, which can tremendously impact their cost and feasibility. \\
\textbf{Results:} Here, we revisit a central ingredient of most in-silico design methods: the sampling of sequences for multi-target design. We establish the $\#${\sf P}-hardness of uniform sampling, and introduce \ourprog{}, a tree decomposition-based algorithm for efficient, fixed parameter tractable sampling. By modeling the problem as a constraint network, \ourprog{} supports generic Boltzmann-weighted sampling for arbitrary additive RNA energy models; this enables generating designs meeting specific goals like expected free energies or \GCb-content. Finally, we empirically study general properties of the approach and generate biologically relevant multi-target Boltzmann-weighted designs for a common design benchmark. In particular, we show significant improvements over seed sequences generated by uniform sampling, the previously best available sampling strategy for multi-target design.\\
\textbf{Availability:} \url{https://github.com/yannponty/RNARedPrint}\\
\textbf{Contact:} Our software is freely available at: \href{yann.ponty@lix.polytechnique.fr}{yann.ponty@lix.polytechnique.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\section{Introduction}
\parHead{Design, applications and motivation for multiple design.}Synthetic biology endeavors the engineering of artificial biological
systems, promising broad applications in biology, biotechnology and
medicine. Centrally, this requires the design of biological
macromolecules with highly specific properties and programmable functions.
In particular, RNAs present themselves as well-suited tools for
rational design targeting specific functions~\cite{Kushwaha2016}. RNA function is tightly
coupled to the formation of secondary structure, as well as changes in
base pairing propensities and the accessibility of regions, e.g. by
burying or exposing interaction sites~\cite{Rodrigo2014}. At the same time, the
thermodynamics of RNA secondary structure is well understood and its prediction is
computationally tractable~\cite{McCaskill1990}. Thus,  structure can serve as effective
proxy within rational design approaches, ultimately targeting catalytic~\cite{Zhang2013} or regulatory~\cite{Rodrigo2014} functions.

\parHead{Motivating multiple RNA design.} The function of many RNAs
depends on their selective folding into one or several alternative
conformations. Classic examples include riboswitches, which
notoriously adopt different stable structures upon binding a specific
ligand. Riboswitches have been a popular application of rational
design~\cite{Wachsmuth2013,Domin2017}, partly motivated by their capacity to
act as biosensors~\cite{Findeiss2017}. At the co-transcriptional
level, certain RNA families feature alternative, evolutionarily
conserved, transient structures~\cite{Zhu2013}, which facilitate the
ultimate adoption of functional structures at full elongation.  More
generally, simultaneous compatibility to multiple structures
is a relevant design objective for engineering kinetically controlled RNAs, finally targeting prescribed folding pathways. Thus,
modern applications of RNA design often target (the simultaneous
stability of) multiple structures, additionally aiming at other
features, such as specific \GCb-content~\cite{Reinharz2013} or the
presence/absence of functionally relevant motifs (either anywhere or
at specific positions)~\cite{Zhou2013}; these objectives motivate flexible
computational design methods.

\parHead{On the importance of sampling for design.}
Many computational methods for RNA design rely on similar overall
strategies: initially generating one or several \Def{seed} sequences
and optimizing them subsequently. In many cases, the seed quality was
found to be critical for the empirical performance (and solution
quality) of RNA design methods~\cite{Levin2012}. For instance, random
seed generation improves the prospect of subsequent optimizations, helping
to overcome local optima of the objective function, and induce better
sequence diversity across designs~\cite{Reinharz2013}.  For
single-target approaches, \Software{INFO-RNA}~\cite{Busch2006} made
significant improvements mainly by starting its local search from the
minimum energy sequence for the target structure instead of (uniform)
random sequences for the early \Software{RNAinverse}
algorithm~\cite{Hofacker1994}. This strategy was later shown to result
in unrealistically high \GCb-contents in designed sequences. To
address this issue, \Software{IncaRNAtion}~\cite{Reinharz2013}
controls the \GCb-content through an adaptive sampling strategy.

\parHead{Specificities and similarities of multi-target design.}
Specifically, for multi-target design, virtually all available methods~\cite{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generation/optimization scheme.
%
Facing the complex sequence constraints induced by multiple targets, early methods such as \Software{Frnakenstein}~\cite{Lyngsoe2012} and \Software{Modena}~\cite{Taneda2015} did not attempt to solve sequence generation systematically, but rely on \emph{ad-hoc} sampling strategies. 
%
Recently, the \Software{RNAdesign} approach~\cite{HoenerzuSiederdissen2013}, coupled with powerful local search within \Software{RNAblueprint}~\cite{Hammer2017}, solved the problem of sampling seeds from the uniform distribution for multiple targets. These methods adopt a graph coloring perspective, initially decomposing the graph hierarchically using various decomposition algorithms, and \Def{precomputing} the number of valid sequences within each subgraph. The decomposition is then reinterpreted as a decision tree to perform a \Def{stochastic backtrack}, inspired by Ding and Lawrence~\cite{Ding2003}, where individual assignments are chosen with probabilities derived from the subsolution counts, leading to uniform sampling. The overall complexity of \Software{RNAdesign} grows like $\Theta(4^{\gamma})$, where $\gamma$ is a parameter whose value is bounded only by the length of the designed RNA.

\parHead{Motivation.} The exponential time and space requirements of the \Software{RNAdesign} method already raise the question of the \Def{complexity of (uniform) sampling for multi-target design}. Since stochastic backtrack can be performed in linear time per sample, the method is dominated by the precomputation step, which requires counting valid designs. Thus, we focus on the question: \emph{Is there a polynomial-time algorithm to count valid multi-target designs?} In Section~\ref{sec:counting}, we answer in the negative, showing that there exists no such algorithm unless ${\sf P}={\sf NP}$. Our result relies on a surprising bijection (up to a trivial symmetry) between valid sequences  and independent sets of a bipartite graph, being the object of recent breakthroughs in approximate counting complexity~\cite{Bulatov2013,Cai2016}.
The hardness of counting (and conjectured hardness of sampling) does not preclude, however, practically applicable algorithms for counting and sampling. In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, in a principled way, from a Boltzmann distribution based on expressive energy models? How to enforce additional constraints, such as the \GCb-content, complex sequence constraints, or the energy of individual structures?}

@@

%In this work, we present a generic approach to multi-target sequence sampling that enables Boltzmann-weighted sampling (subsuming uniform sampling) for a wide class of objective functions that can even make use of the commonly used ``full-featured'' RNA energy models (like the Turner model~\cite{turner}). Due to its high expressivity the same framework allows combining objective functions with various hard and soft constraints, which can accommodate various desirable multi-target design scenarios.

%To provide a simple concrete example of multi-target design, consider designing a single sequence where all target structures are as stable as possible.
%That is, one asks for the sequence $S$ that minimizes the energy differences of the target structures to the minimum energy for this sequence among all possible structures; one aims to minimize
%   \begin{equation}
%\label{eq:design-objective}
%\sum_{\ell\in[1,k]} E(S;R_\ell) - E_0(S),
%\end{equation}
%over all sequences $S$, where $E_0(S):=\min_R E(S;R)$.
%
%The proposed Boltzmann-weighted sampling approach can support this design procedure by sampling start structures from the Boltzmann distribution over
%\begin{equation}
%\label{eq:sampling-objective} \sum_{\ell\in[1,k]} w_\ell E(S;R_\ell)
%\end{equation}
%or extensions thereof, e.g.~for controlling base frequencies. Suitable weights $w_\ell$ to accommodate the design objective can be approximated automatically after the machinery for Boltzmann sampling over such functions is set up.
%
%\TODO{Rewrite (from here to "Contributions"): sounds too negative}
%Notably, by Eq.~(\ref{eq:sampling-objective}) we do not suggest to sample over the energy difference of Eq.~$(\ref{eq:design-objective})$. The term $\min_R E(S;R)$, which---for each sequence---minimizes over the entire structure space, precludes efficient sampling even by the proposed approach.
%
%Already sampling over Eq.~(\ref{eq:sampling-objective}) strictly extends the only previously existing approach to controlled sampling from single to multi-target sampling (in complete analogy to IncaRNAtion only after GC content is controlled, which we present later) and subsumes uniform sampling for multi-target design.
%
%Naturally, also more complex objective functions are desirable for RNA design and have been discussed in the literature; e.g. the ensemble defect~\cite{Zadeh:2010}. Typically, more complex objective functions make design computationally much more challenging. Similarily, also the proposed sampling framework supports only a limited class of objective functions directly and efficiently---namely ``only'' all additive energy functions. Nevertheless, one can safely assume that the design under various more realistic objective functions profits from controlled sampling as long as the sampling is directed towards the final objective function.

@@

%We solve the problem efficiently for any fixed parameter $w$, whose value reflects the intrinsic complexity of the structures and energy function $E_\R$.
%
As will be explicited in subsequent sections, our approach relies on breaking down the energy function $E_\R(S)$ into additive components, each depending on only few sequence positions. Given $\R$, we express $E_\R(S)$
as the sum of energy contributions $f(S)$ over a set $\F$ (of
functions $f:\S_n\to\real$), s.t.~$E_\R(S)=\sum_{f\in\F} f(S)$. This
captures realistic RNA energy models, while bounding the
dependencies to sequence positions introduced by each single $f$. Formally, define the \Def{dependencies} $\dep(f)$ of $f$ as a minimal set of sequence positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all sequences $S$ and $S'$ agreeing at all positions in $\I$.

@@

%%% reformulate later
%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$. Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$. As previously noted~\cite{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides. The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.

@@

We remind that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$. Let $\IS{G}$ denote the set of all independent sets in $G$. 
Now, consider the function $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$

\begin{proposition}
 $\Psi$ is a bijection between $\Design{G}$ and $\IS{G}$.
\end{proposition}
\begin{proof}
Let us establish the injectivity of  $\Psi$, i.e. that $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
If $f\neq f'$, then there exists a node $v\in V$ such that $f(v)\neq f'(v)$. 
Assume that $v\in V_1$, and remind that the only nucleotides allowed in $V_1$ are $\Ab$ and $\Gb$. Since $f(v)\neq f'(v)$, then we have $\{f(v),f'(v)\}=\{\Ab,\Gb\}$
and we conclude that $\Psi(f)$ differs from $\Psi(f')$ at least with respect to its inclusion/exclusion of $v$.

We turn to the surjectivity of $\Psi$, i.e. the existence of a preimage for each element $S\in \IS{G}$. Let us consider the function $f$ defined as
\begin{align}
 \forall v_1\in V_1, f(v_1)&= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases}& 
 \forall v_2\in V_2, f(v_2)&= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
\end{align}
One easily verifies that $\Psi(f) = S$.

@@

Now we can build on the connection between the two problems to obtain complexity results for \NumDesign. Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\cite{Ge2012}, from which we immediately conclude:
\begin{corollary}
  \NumDesign is $\#{\sf P}$-hard
\end{corollary}
\begin{proof}
  Note that $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ is given by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$. Thus any efficient algorithm for $\#{\sf BIS}$ on connected instances provides an efficient algorithm for general graphs.

@@

\section{An FPT algorithm for sampling Boltzmann-weighted designs}
\label{sec:FPT}

For our algorithmic description, we translate the concepts of
Section~\ref{sec:problem-statement} to the formalism of constraint networks, here
specialized as RNA design network. This allows us to base our
algorithm on cluster tree elimination (CTE) by Rina
Dechter~\cite{Dechter2013}.
%
In the RNA design network, sequences $S\in\S_n$ are isomorphically
modeled as assignments $\val_S$ of bases to $n$ variables, each associated to
a sequence position. The functions $f\in\F$ of
Section~\ref{sec:problem-statement}, are interpreted as functions on the network's variables.

@@

Let us now consider the \Def{computation of the partition function}.
Given is the RNA design network $\network=(\X,\F)$ and its cluster
tree decomposition $(T,\chi,\phi)$.  W.l.o.g, we assume that $T$ is
connected and contains a dedicated node $r$, with $\chi(r)=\varnothing$
and $\phi(r)=\varnothing$, added as a virtual root
$r$ connected to a node in each connected
component of $T$.  Now, we consider the set of directed edges
$\edgesToR{}$ of $T$ oriented to $r$; define $T_r(u)$ as the induced
subtree of $u$. Algorithm~\ref{alg:pf} computes the partition function by
passing messages along these directed edges $u\to v$ (i.e. always from
some child $u$ to its parent $v$). Each message is a function on
variables $\vars(m)\subseteq \X$ to $\real\cup\{\infty\}$. The message
from $u$ to $v$ represents the partition functions of the subtree of
$u$ for all possible assignments to the variables
$\vars(m)=\separator{u}{v}$. The correctness of the algorithm can be shown by induction over $T$ (Appendix~\ref{appsec:correctness}).
After running Alg.~\ref{alg:pf}, 
multiplying the 0-ary messages sent to the root $r$ yields the total partition function:
\begin{math}
  \partfun{\network} = \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

The partition functions can then direct a \Def{stochastic backtrack} to achieve \Def{Boltzmann sampling of sequences}, such that one samples from the Boltzmann distribution of a given desgin network $\network$. The sampling algorithm assumes that the cluster tree was expanded and the messages $\Message{u}{v}$ for the edges in $\edgesToR{}$ are already generated by Algorithm \ref{alg:pf} for the expanded cluster tree.
%
Algorithm~\ref{alg:sampling} defines the recursive procedure $\sample(u,\val)$, which returns an  assignment of all variables in the subtree rooted at $u$ from the Boltzmann-distribution.
Called on $r$ and the empty assignment, the procedure returns a total assignment corresponding to a Boltzmann-distributed random sequence.

@@

First, we observed that the computation time of tree decomposition
(\Software{GreedyFillIn}, implemented in \Software{LibTW}~\cite{Dijk2006}) for multi-target
sampling is negligible compared to Alg.~\ref{alg:pf} (Appendix~\ref{appsec:treedecomp} and \ref{appsec:dependency-cliques});
consequently, we discuss complexities without the tree decomposition
step.

For our analysis, we define the \emph{maximum separator size} $s$ as
$\max_{u,v\in V} | \separator{u}{v} |$ and denote the maximum size of
$\difference{u}{v}$ over $(u,v)\in\edgesToR{}$ as $D$.  Generally,
without specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and $\mathcal{O}(|V|\cdot4^s)$ space (Appendix~\ref{appsec:algcomplexity});
Alg.~\ref{alg:sampling} would require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per
sample on arbitrary tree decompositions (Appendix~\ref{appsec:algcomplexity}). W.l.o.g. we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.

@@

As already noted by Flamm \emph{et al}~\cite{Flamm2001}, the
requirement of canonical base pairing induces a bi-partition of each connected
component of the compatibility graph, such that the variables of one
partition are restricted to values $\{\Ab,\Gb\}$, but for other
partition, to values $\{\Cb,\Ub\}$. This suggests a strong
optimization of our algorithms, which reduces the basis in the
exponential factor from 4 to 2. The price for this reduction is a
binary choice for each of $c$ connected component. For each of the
$2^c$ combinations, the partition functions can be computed separately
in $O(n\,m\, 2^{w+1})$ time. This strategy is applied only if
$c< w+1$. 
%% SW: This could be optimized further by applying the strategy only to (optimally) selected components
Finally, we compute the sums of the partition functions for all possible combinations.

>>>>

\abstract{\textbf{Motivation:} The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches largely benefit from effective in-silico methods, which can tremendously impact their cost and feasibility. \\
\textbf{Results:} Here, we revisit a central ingredient of most in-silico design methods: the sampling of sequences for multi-target design. We establish the $\#${\sf P}-hardness of uniform sampling, and introduce \ourprog{}, a tree decomposition-based algorithm for efficient, fixed parameter tractable sampling. By modeling the problem as a constraint network, \ourprog{} supports generic Boltzmann-weighted sampling for arbitrary additive RNA energy models; this enables generating designs meeting specific goals like expected free energies or \GCb-content. Finally, we empirically study general properties of the approach and generate biologically relevant multi-target Boltzmann-weighted designs for a common design benchmark. In particular, we show significant improvements over seed sequences generated by uniform sampling, the previously best available sampling strategy for multi-target design.\\
\textbf{Availability:} Our software is freely available at: \url{https://github.com/yannponty/RNARedPrint}\\
\textbf{Contact:} \href{yann.ponty@lix.polytechnique.fr}{yann.ponty@lix.polytechnique.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\section{Introduction}
\parHead{Design, applications and motivation for multiple design.}Synthetic biology endeavors the engineering of artificial biological
systems, promising broad applications in biology, biotechnology and
medicine. Centrally, this requires the design of biological
macromolecules with highly specific properties and programmable functions.
In particular, RNAs present themselves as well-suited tools for
rational design targeting specific functions~\citep{Kushwaha2016}. RNA function is tightly
coupled to the formation of secondary structure, as well as changes in
base pairing propensities and the accessibility of regions, e.g. by
burying or exposing interaction sites~\citep{Rodrigo2014}. At the same time, the
thermodynamics of RNA secondary structure is well understood and its prediction is
computationally tractable~\citep{McCaskill1990}. Thus,  structure can serve as effective
proxy within rational design approaches, ultimately targeting catalytic~\citep{Zhang2013} or regulatory~\citep{Rodrigo2014} functions.

\parHead{Motivating multiple RNA design.} The function of many RNAs
depends on their selective folding into one or several alternative
conformations. Classic examples include riboswitches, which
notoriously adopt different stable structures upon binding a specific
ligand. Riboswitches have been a popular application of rational
design~\citep{Wachsmuth2013,Domin2017}, partly motivated by their capacity to
act as biosensors~\citep{Findeiss2017}. At the co-transcriptional
level, certain RNA families feature alternative, evolutionarily
conserved, transient structures~\citep{Zhu2013}, which facilitate the
ultimate adoption of functional structures at full elongation.  More
generally, simultaneous compatibility to multiple structures
is a relevant design objective for engineering kinetically controlled RNAs, finally targeting prescribed folding pathways. Thus,
modern applications of RNA design often target (the simultaneous
stability of) multiple structures, additionally aiming at other
features, such as specific \GCb-content~\citep{Reinharz2013} or the
presence/absence of functionally relevant motifs (either anywhere or
at specific positions)~\citep{Zhou2013}; these objectives motivate flexible
computational design methods.

\parHead{On the importance of sampling for design.}
Many computational methods for RNA design rely on similar overall
strategies: initially generating one or several \Def{seed} sequences
and optimizing them subsequently. In many cases, the seed quality was
found to be critical for the empirical performance (and solution
quality) of RNA design methods~\citep{Levin2012}. For instance, random
seed generation improves the prospect of subsequent optimizations, helping
to overcome local optima of the objective function, and induce better
sequence diversity across designs~\citep{Reinharz2013}.  For
single-target approaches, \Software{INFO-RNA}~\citep{Busch2006} made
significant improvements mainly by starting its local search from the
minimum energy sequence for the target structure instead of (uniform)
random sequences for the early \Software{RNAinverse}
algorithm~\citep{Hofacker1994}. This strategy was later shown to result
in unrealistically high \GCb-contents in designed sequences. To
address this issue, \Software{IncaRNAtion}~\citep{Reinharz2013}
controls the \GCb-content through an adaptive sampling strategy.

\parHead{Specificities and similarities of multi-target design.}
Specifically, for multi-target design, virtually all available methods~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generation/optimization scheme.
%
Facing the complex sequence constraints induced by multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} did not attempt to solve sequence generation systematically, but rely on \emph{ad-hoc} sampling strategies. 
%
Recently, the \Software{RNAdesign} approach~\citep{HoenerzuSiederdissen2013}, coupled with powerful local search within \Software{RNAblueprint}~\citep{Hammer2017}, solved the problem of sampling seeds from the uniform distribution for multiple targets. These methods adopt a graph coloring perspective, initially decomposing the graph hierarchically using various decomposition algorithms, and \Def{precomputing} the number of valid sequences within each subgraph. The decomposition is then reinterpreted as a decision tree to perform a \Def{stochastic backtrack}, inspired by Ding and Lawrence~\citep{Ding2003}, where individual assignments are chosen with probabilities derived from the subsolution counts, leading to uniform sampling. The overall complexity of \Software{RNAdesign} grows like $\Theta(4^{\gamma})$, where $\gamma$ is a parameter whose value is bounded only by the length of the designed RNA.

\parHead{Motivation.} The exponential time and space requirements of the \Software{RNAdesign} method already raise the question of the \Def{complexity of (uniform) sampling for multi-target design}. Since stochastic backtrack can be performed in linear time per sample, the method is dominated by the precomputation step, which requires counting valid designs. Thus, we focus on the question: \emph{Is there a polynomial-time algorithm to count valid multi-target designs?} In Section~\ref{sec:counting}, we answer in the negative, showing that there exists no such algorithm unless ${\sf P}={\sf NP}$. Our result relies on a surprising bijection (up to a trivial symmetry) between valid sequences  and independent sets of a bipartite graph, being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.
The hardness of counting (and conjectured hardness of sampling) does not preclude, however, practically applicable algorithms for counting and sampling. In particular, we wish to extend the flexibility of multi-structure design, leading to the following questions: \emph{How to sample, in a principled way, from a Boltzmann distribution based on expressive energy models? How to enforce additional constraints, such as the \GCb-content, complex sequence constraints, or the energy of individual structures?}

@@

%In this work, we present a generic approach to multi-target sequence sampling that enables Boltzmann-weighted sampling (subsuming uniform sampling) for a wide class of objective functions that can even make use of the commonly used ``full-featured'' RNA energy models (like the Turner model~\citep{turner}). Due to its high expressivity the same framework allows combining objective functions with various hard and soft constraints, which can accommodate various desirable multi-target design scenarios.

%To provide a simple concrete example of multi-target design, consider designing a single sequence where all target structures are as stable as possible.
%That is, one asks for the sequence $S$ that minimizes the energy differences of the target structures to the minimum energy for this sequence among all possible structures; one aims to minimize
%   \begin{equation}
%\label{eq:design-objective}
%\sum_{\ell\in[1,k]} E(S;R_\ell) - E_0(S),
%\end{equation}
%over all sequences $S$, where $E_0(S):=\min_R E(S;R)$.
%
%The proposed Boltzmann-weighted sampling approach can support this design procedure by sampling start structures from the Boltzmann distribution over
%\begin{equation}
%\label{eq:sampling-objective} \sum_{\ell\in[1,k]} w_\ell E(S;R_\ell)
%\end{equation}
%or extensions thereof, e.g.~for controlling base frequencies. Suitable weights $w_\ell$ to accommodate the design objective can be approximated automatically after the machinery for Boltzmann sampling over such functions is set up.
%
%\TODO{Rewrite (from here to "Contributions"): sounds too negative}
%Notably, by Eq.~(\ref{eq:sampling-objective}) we do not suggest to sample over the energy difference of Eq.~$(\ref{eq:design-objective})$. The term $\min_R E(S;R)$, which---for each sequence---minimizes over the entire structure space, precludes efficient sampling even by the proposed approach.
%
%Already sampling over Eq.~(\ref{eq:sampling-objective}) strictly extends the only previously existing approach to controlled sampling from single to multi-target sampling (in complete analogy to IncaRNAtion only after GC content is controlled, which we present later) and subsumes uniform sampling for multi-target design.
%
%Naturally, also more complex objective functions are desirable for RNA design and have been discussed in the literature; e.g. the ensemble defect~\citep{Zadeh:2010}. Typically, more complex objective functions make design computationally much more challenging. Similarily, also the proposed sampling framework supports only a limited class of objective functions directly and efficiently---namely ``only'' all additive energy functions. Nevertheless, one can safely assume that the design under various more realistic objective functions profits from controlled sampling as long as the sampling is directed towards the final objective function.

@@

%We solve the problem efficiently for any fixed parameter $w$, whose value reflects the intrinsic complexity of the structures and energy function $E_\R$.
%
As will be explicit in subsequent sections, our approach relies on breaking down the energy function $E_\R(S)$ into additive components, each depending on only few sequence positions. Given $\R$, we express $E_\R(S)$
as the sum of energy contributions $f(S)$ over a set $\F$ (of
functions $f:\S_n\to\real$), s.t.~$E_\R(S)=\sum_{f\in\F} f(S)$. This
captures realistic RNA energy models, while bounding the
dependencies to sequence positions introduced by each single $f$. Formally, define the \Def{dependencies} $\dep(f)$ of $f$ as a minimal set of sequence positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all sequences $S$ and $S'$ agreeing at all positions in $\I$.

@@

%%% reformulate later
%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$. Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$. As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides. The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.

@@

We remind that an \Def{independent set} of $G=(V,E)$ is a subset $V'\subseteq V$ of nodes that are not connected by any edge in $E$. Let $\IS{G}$ denote the set of all independent sets in $G$.

\begin{proposition}
 $\Design{G}$ is in bijection with $\IS{G}$.
\end{proposition}
\begin{proof}
Consider the function $\Psi: \Design{G} \to \IS{G}$ defined by $ \Psi(f) := \left\{v\in V\mid f(v)\in\{\Ab,\Cb\}\right\}.$

Let us establish the injectivity of  $\Psi$, i.e. that $\Psi(f)\neq\Psi(f')$ for all $f\neq f'$.
If $f\neq f'$, then there exists a node $v\in V$ such that $f(v)\neq f'(v)$. 
Assume that $v\in V_1$, and remind that the only nucleotides allowed in $V_1$ are $\Ab$ and $\Gb$. Since $f(v)\neq f'(v)$, then we have $\{f(v),f'(v)\}=\{\Ab,\Gb\}$
and we conclude that $\Psi(f)$ differs from $\Psi(f')$ at least with respect to its inclusion/exclusion of $v$.

We turn to the surjectivity of $\Psi$, i.e. the existence of a preimage for each element $S\in \IS{G}$. Let us consider the function $f$ defined as
\begin{align}
 \forall v_1\in V_1, f(v_1)&= \begin{cases} \Ab & \text{if }v_1\in S\\ \Gb & \text{if }v_1\notin S\end{cases}\\
 \forall v_2\in V_2, f(v_2)&= \begin{cases} \Cb & \text{if }v_2\in S\\ \Ub & \text{if }v_2\notin S\end{cases}
\end{align}
One easily verifies that $\Psi(f) = S$.

@@

Now we can build on the connection between the two problems to obtain complexity results for \NumDesign. Counting independent sets in bipartite graphs ($\#{\sf BIS}$) is indeed a well-studied \#{\sf P}-hard problem~\citep{Ge2012}, from which we immediately conclude:
\begin{corollary}
  \NumDesign is $\#{\sf P}$-hard
\end{corollary}
\begin{proof}
  Note that $\#{\sf BIS}$ is also \#{\sf P}-hard on connected graphs, as the number of independent sets for a disconnected graph $G$ is given by $|\IS{G}|=\prod_{cc\in CC(G)} |\IS{cc}|$. Thus any efficient algorithm for $\#{\sf BIS}$ on connected instances provides an efficient algorithm for general graphs.

@@

\section{An FPT algorithm for the partition function and sampling of Boltzmann-weighted designs}
\label{sec:FPT}

For our algorithmic description, we translate the concepts of
Section~\ref{sec:problem-statement} to the formalism of constraint networks, here
specialized as RNA design network. This allows us to base our
algorithm on cluster tree elimination (CTE) by Rina
Dechter~\citep{Dechter2013}.
%
In the RNA design network, sequences $S\in\S_n$ are isomorphically
modeled as assignments $\val_S$ of bases to $n$ variables, each associated to
a sequence position. The functions $f\in\F$ of
Section~\ref{sec:problem-statement}, are interpreted as functions on the network's variables.

@@

Let us now consider the \Def{computation of the partition function}.
Given is the RNA design network $\network=(\X,\F)$ and its cluster
tree decomposition $(T,\chi,\phi)$.  W.l.o.g, we assume that $T$ is
connected and contains a dedicated node $r$, with $\chi(r)=\varnothing$
and $\phi(r)=\varnothing$, added as a virtual root
$r$ connected to a node in each connected
component of $T$.  Now, we consider the set of directed edges
$\edgesToR{}$ of $T$ oriented to $r$; define $T_r(u)$ as the induced
subtree of $u$. Algorithm~\ref{alg:pf} computes the partition function by
passing messages along these directed edges $u\to v$ (i.e. always from
some child $u$ to its parent $v$). Each message is a function on
variables $\vars(m)\subseteq \X$ to $\real\cup\{\infty\}$. The message
from $u$ to $v$ represents the partition functions of the subtree of
$u$ for all possible assignments to the variables
$\vars(m)=\separator{u}{v}$. The correctness of the algorithm can be shown by induction over $T$ (Supp. Mat.~\ref{appsec:correctness}).
After running Alg.~\ref{alg:pf}, 
multiplying the 0-ary messages sent to the root $r$ yields the total partition function:
\begin{math}
  \partfun{\network} = \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

The partition functions can then direct a \Def{stochastic backtrack} to achieve \Def{Boltzmann sampling of sequences}, such that one samples from the Boltzmann distribution of a given design network $\network$. The sampling algorithm assumes that the cluster tree was expanded and the messages $\Message{u}{v}$ for the edges in $\edgesToR{}$ are already generated by Algorithm \ref{alg:pf} for the expanded cluster tree.
%
Algorithm~\ref{alg:sampling} defines the recursive procedure $\sample(u,\val)$, which returns an  assignment of all variables in the subtree rooted at $u$ from the Boltzmann-distribution.
Called on $r$ and the empty assignment, the procedure returns a total assignment corresponding to a Boltzmann-distributed random sequence.

@@

First, we observed that the computation time of tree decomposition
(\Software{GreedyFillIn}, implemented in \Software{LibTW}~\citep{Dijk2006}) for multi-target
sampling is negligible compared to Alg.~\ref{alg:pf} (Supp. Mat.~\ref{appsec:treedecomp} and \ref{appsec:dependency-cliques});
we will thus omit the tree decomposition from our complexity analysis.

For our analysis, we define the \emph{maximum separator size} $s$ as
$\max_{u,v\in V} | \separator{u}{v} |$ and denote the maximum size of
$\difference{u}{v}$ over $(u,v)\in\edgesToR{}$ as $D$.  In the absence
of specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and $\mathcal{O}(|V|\cdot4^s)$ space (Supp. Mat.~\ref{appsec:algcomplexity});
Alg.~\ref{alg:sampling} would require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per
sample on arbitrary tree decompositions (Supp. Mat.~\ref{appsec:algcomplexity}). W.l.o.g. we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.

@@

As already noted by \citet{Flamm2001}, the
requirement of canonical base pairing induces a bi-partition of each connected
component of the compatibility graph, such that the variables of one
partition are restricted to values $\{\Ab,\Gb\}$, but for other
partition, to values $\{\Cb,\Ub\}$. This suggests a strong
optimization of our algorithms, which reduces the basis in the
exponential factor from 4 to 2. The price for this reduction is a
binary choice for each of $c$ connected component. For each of the
$2^c$ combinations, the partition functions can be computed separately
in $O(n\,m\, 2^{w+1})$ time. This strategy is applied only if
$c< w+1$. 
%% SW: This could be optimized further by applying the strategy only to (optimally) selected components
Finally, we compute the sums of the partition functions for all possible combinations.
